Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-01 numerical-gravity.tex

Solving for the accelerations of the cells and particles on the grid due to self-gravity involves three steps: (i) computing the total gravitating mass, (ii) solving for the potential field with the appropriate boundary conditions, and (iii) differencing the potential to get the acceleration, and, if necessary, interpolating the potential back to the particles. These steps are described in detail below.

First, the dark matter particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citet{Hockney88} to form a spatially discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids which lie within the entire gravitating field are used.  This step can involve communication.

Next, we add the baryonic grid densities in a similar fashion.  In particular, we treat baryonic cells as virtual CIC particles which are are placed at the grid center but are advanced by $0.5 \Delta t v^n$ in order to approximately time-center the gravitating mass field.  Cells which are covered by further refined grids are treated in a similar way (i.e. we also use the sub grid cells as virtual CIC particles).  This procedure results in a total gravitating mass field $\rho_{\rm total}^{n+1/2}$.

@@

For isolated boundary conditions, we use James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In either case, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference to obtain accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

In order to calculate more accurate potentials on subgrids, \enzo\ uses a similar but slightly different technique on the root grids.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub grids as before.  To compute the potential on sub grids; however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each sub grid separately.

The region immediately next to the boundary can contain unwanted oscillations \citep[e.g.,][]{Anninos94}, and so we use an expanded buffer zone around the grid, of size three root grid boundary zones (so typically six refined zones for a refinement factor of 2).  The density is computed in this region and the potential solved, but only the region which overlaps with the active region of grid itself is used to calculate accelerations.

It is known (REF) that simply interpolating the potential without feeding it back to higher levels leads to errors in the potential at more refined levels because of the build-up of errors during the interpolation of coarse boundary values.  In addition, neighboring subgrids are not guaranteed to generate the same potential values because of the lack of a coherent potential solve across the whole hierarchy. In an attempt to partially alleviate this problem, we allow for an iterative procedure across sibling grids, in which the potential values on the boundary of grids can be updated with the potential in `active' regions of neighboring subgrids.  To prevent overshoot, we average the potential on the boundary and allow for (typically) 4 iterations.  This procedure can help in many cases, but does not necessarily produce a coherent solution across all grids and so does not completely solve the problem; we are working on a slower but more accurate method which does a multi-grid solve across the whole grid (Reynolds \etal, in preparation).

>>>>

Solving for the accelerations of the cells and particles on the grid due to self-gravity involves three steps: (i) computing the total gravitating mass, (ii) solving for the gravitational potential field with the appropriate boundary conditions, and (iii) differencing the potential to get the acceleration, and, if necessary, interpolating the acceleration back to the particles. These steps are described in detail below.

First, the dark matter particles are distributed onto the grids using the second-order cloud-in-cell (CIC) interpolation technique \citep{Hockney88} to form a spatially-discretized density field $\rho_{\rm DM}$.  When doing the CIC interpolation, particle positions are (temporarily) advanced by $0.5 \Delta t v^n$ so that we generate an estimate of the time-centered density field.  Particles on sub-grids are also added in this way.  In addition, since the gravitating field for a grid is defined beyond the grid edges (see below), dark matter particles from sibling grids and sub grids whhatich lie within the entire gravitating field are used.  This step can involve communication.

Next, we add the baryonic grid densities in a similar fashion.  In particular, we treat baryonic cells as virtual CIC particles that are are placed at the grid center but are advanced by $0.5 \Delta t v^n$ in order to approximately time-center the gravitating mass field.  Cells that are covered by further-refined grids are treated in a similar way (i.e., we also use the subgrid cells as virtual CIC particles).  This procedure results in a total gravitating mass field $\rho_{\rm total}^{n+1/2}$.

@@

For isolated boundary conditions, we use the James method \citep{James77}.  In this case, the Greens function is generated in real-space so as to have the correct zero-padding properties and then transformed into the Fourier domain.  In either case, the potential is then transformed back into the real domain to get potential values at the cell centers.  These are differenced with a two-point centered difference scheme to obtain accelerations at the cell centers (except if we are using the staggered Zeus-like solver, in which case the accelerations are computed at the cell faces to match the velocities).  Particle accelerations are obtained using a (linear) CIC interpolation from the grid.

In order to calculate more accurate potentials on subgrids, \enzo\ uses a similar but slightly different technique from the root grid.  The generation of the total gravitating mass field is essentially identical, using CIC interpolation for both the particles and baryons, including sub grids as before.  To compute the potential on sub grids, however, we use the standard seven-point (in three dimensions) second-order finite-difference approximation to Poisson's equation.  Boundary conditions are then interpolated from the potential values on the parent grid.  We use either tri-linear interpolation or a natural second-order spline for this: both methods give similar results. The potential equation on each subgrid is then solved with the given Dirichlet boundary conditions with a multigrid relaxation technique.  This is applied to each sub grid separately.

The region immediately next to the boundary can contain unwanted oscillations \citep[e.g.,][]{Anninos94}, and so we use an expanded buffer zone around the grid, of size three parent grid boundary zones (so typically six refined zones for a refinement factor of 2).  The density is computed in this region and the potential solved, but only the region that overlaps with the active region of the grid itself is used to calculate accelerations.

It is known \red{(REF)} that simply interpolating the potential without feeding it back to higher levels leads to errors in the potential at more refined levels due to the build-up of errors during the interpolation of coarse boundary values.  In addition, neighboring subgrids are not guaranteed to generate the same potential values because of the lack of a coherent potential solve across the whole hierarchy. In an attempt to partially alleviate this problem, we allow for an iterative procedure across sibling grids, in which the potential values on the boundary of grids can be updated with the potential in `active' regions of neighboring subgrids.  To prevent overshoot, we average the potential on the boundary and allow for (typically) 4 iterations.  This procedure can help in many cases, but does not necessarily produce a coherent solution across all grids and so does not completely solve the problem; we are working on a slower but more accurate method that does a multi-grid solve across the whole grid (Reynolds \etal, in preparation).

### 2013-05-01 numerical-ppm.tex

We can examine the structure of eqs.~(\ref{eq:mass})-(\ref{eq:total_energy})
to obtain a better understanding of the effect of each term.
The derivatives are first order in time
and provide us with the dependent variables that will be evaluated each
time step: $\rhob$, $\vecvb$, and $E$.

@@

The first term on the right hand side of eqs. (\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in eq.~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  It does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.

@@

The remainder of the terms (transport and source) in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
Once again we use the concept of operator splitting,
but now apply it spatially.   We can rewrite the one-dimensional (Eulerian) 
versions of eqs.~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}

%
Note that $x$ and $v$ refer to the one dimensional comoving position and 
peculiar velocity of the baryonic gas, and $g$ is the acceleration at the cell center.
We write this including the cosmological scale-factors but it reduces to the non-cosmological case with $a = 1$.
These equations are now in a form that can be solved by the split PPM scheme.

@@

We now restrict ourselves to the solution of eqs. (\ref{eq:mass1d})--(\ref{eq:total_energy1d}), plus transport, in one dimension.  In the Direct Eulerian version of PPM, this is accomplished by a three-step process.  First, we compute effective left and right states at each grid boundary by constructing a piecewise parabolic description of the primary variables ($\rho$, $u$ and $E$) and then averaging over the regions corresponding to the distance each characteristic wave can travel ($u$, $u-c$ and $u+c$).  Second, a Riemann problem is solved using these effective left and right states, and finally the fluxes are computed based on the solution to this Riemann problem.  This is described in detail in \citet{1984JCoPh..54..174C}, but we will briefly outline the procedure here both for completeness and to put the changes we will make in context.

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$  The pressure is determined from
eq.~(\ref{eq:state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is eq.~(1.4) of CW84 (in the radial rather than mass coordinate as is appropriate for the direct rulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see eqs. 1.7 to 1.10 of \citet{1984JCoPh..54..174C}.
We also optionally allow steepening as described in this reference.

@@

In the Eulerian approach, this is more complex because the Eulerian zone-boundary does not naturally correspond to one the three characteristic waves and so a more complicated approach is required.  First, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{eqnarray}
      q^{C+}_{j+1/2} & = & f_{j+1/2,L}(\Delta t C^n_j), \\
      q^{C-}_{j+1/2} & = & f_{j+1/2,R}(-\Delta t C^n_{j+1}), \nonumber
\end{eqnarray}
where $q^{C+}$ ($q^{C-}$) refers to the left (right) characteristic $C$.  The functions $f$ are just integrals
over the parabolic interpolation formula:
%
\begin{eqnarray}
f_{j+1/2,L}(y) & = &
       {1 \over y} \int^{x_{j+1/2}}_{x_{j+1/2}-y} q_j(x) dx, \\
f_{j+1/2,R}(y) & = &
       {1 \over y} \int^{x_{j+1/2}+y}_{x_{j+1/2}} q_{j+1}(x) dx. \nonumber
\end{eqnarray}
The linearized gas dynamics equations are then used to compute second-order correct left and right states which take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
with an iterative approach \citep[see][]{Woodward86}, producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then used to solve the difference eqs. (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

We include an optional diffusive flux (and flattening for the parabolic curves) which can improve the solution in some cases.  Our implementation follows that in the appendix of \citet{1984JCoPh..54..174C}.

In addition, as discussed earlier, the three dimensional scheme is achieved by operator splitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see eq. 3.6 of \citet{1984JCoPh..54..174C}).

We note that the acceleration required in eq.~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

The system described so far works well for gravitating systems with
reasonable Mach numbers ($<100$) as long as the structures are well resolved.
This section and the next detail changes that are required
to correctly account for situations in which one or both of these
requirements are not met.

Large, hypersonic bulk flows appear to be very common in cosmological
simulations and they present a problem because of the high ratio of
kinetic energy $E_k$ to gas internal energy $e$, which can reach as
high as $10^8$.  Inverted, we see that the internal energy consists of
an extremely small portion of the total energy.  The pressure then,
proportional to $E - E_{k}$, is the small difference between two large
numbers: a disastrous numerical situation.  This is not as large a
problem as it may at first appear because it only occurs when the
pressure is negligibly small.  Therefore, even if we suffer large
errors in the pressure distribution in these regions, the dynamics and
total energy budget of the flow will remain unaffected.  Nevertheless,
if the temperature distribution is required for other reasons
(e.g. radiative processes), a remedy is required.

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \nabla e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \nabla \cdot \vecv
\end{equation}
in comoving co-ordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, is not in flux-conservative form.  In particular, the pressure is evaluated at the cell-center.  Unfortunately, time-centering of this pressure has proved difficult to do without generating large errors in the internal energy and so we conservatively leave the pressure at the old time in this difference equation.  This leads to spreading of shocks; however, we note that this equation is only used in hypersonic flows.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$ which is consistent with the truncation error of the
scheme.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.08$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools to the expansion of the universe
but large bulk flows develop due to the formation of structure.

>>>>

We can examine the structure of Equations~(\ref{eq:mass})-(\ref{eq:total_energy})
to obtain a better understanding of the effect of each term.
The derivatives are first order in time
and provide us with the dependent variables that will be evaluated each
time step: $\rhob$, $\vecvb$, and $E$.

@@

The first term on the right hand side of Equations~(\ref{eq:momentum}) and
(\ref{eq:total_energy}) comes from our choice of comoving coordinates
(a similar term also appears in Equation~(\ref{eq:dm_velocity}), the velocity 
relation for the dark matter particles).  It does not
appear in the mass conservation equation (\ref{eq:mass}) because
of the comoving density definition.   
We note that these expansion terms could be
eliminated entirely by the proper choice of variables (including time),
although we have not done so here, as they do not constitute a major
source of error.

@@

The remainder of the terms (transport and source) in the fluid equations
involve derivatives.  
Because we are interested in phenomena with no special geometry, we will
restrict discussion to Cartesian coordinates.
Once again we use the concept of operator splitting,
but now apply it spatially.   We can rewrite the one-dimensional (Eulerian) 
versions of Equations~(\ref{eq:mass})--(\ref{eq:total_energy}) without expansion terms
in conservative form as,
\begin{eqnarray}
\frac{\partial \rho}{\partial t}  + \frac{1}{a} \frac{\partial \rho v }{\partial x}    & = &  
     0 \label{eq:mass1d} \\
\frac{\partial \rho v}{\partial t}  + \frac{1}{a} \frac{\partial \rho v^2}{\partial x}   + 
      \frac{1}{a} \frac{\partial p}{ \partial x} & = & 
      \rho \frac{1}{a} \frac{\partial \phi}{ \partial x}  \label{eq:momentum1d}  \\
\frac{\partial \rho E}{\partial t}  + \frac{1}{a} \frac{\partial \rho v E}{\partial x}  & =  &
      \rho v \frac{1}{a} \frac{\partial \phi}{\partial x}. \label{eq:total_energy1d}
\end{eqnarray}

%
Note that $x$ and $v$ refer to the one dimensional comoving position and 
peculiar velocity of the baryonic gas, and $g$ is the acceleration at the cell center.
We write this including the cosmological scale factors, but it reduces to the non-cosmological case with $a = 1$.
These equations are now in a form that can be solved by the split PPM scheme.

@@

We now restrict ourselves to the solution of Equations
(\ref{eq:mass1d})--(\ref{eq:total_energy1d}), plus transport, in one
dimension.  In the Direct Eulerian version of PPM, this is
accomplished by a three-step process.  First, we compute effective
left and right states at each grid boundary by constructing a
piecewise parabolic description of the primary variables ($\rho$, $u$
and $E$) and then averaging over the regions corresponding to the
distance each characteristic wave can travel ($u$, $u-c$ and $u+c$).
Second, a Riemann problem is solved using these effective left and
right states, and finally the fluxes are computed based on the
solution to this Riemann problem.  This is described in detail in
\citet{1984JCoPh..54..174C}, but we will briefly outline the procedure
here both for completeness and to put the changes we will make in
context.

@@

We first construct monotonic piecewise 
parabolic (third order) interpolations in one dimension
for each of $p$, $\rho$, and $v$  The pressure is determined from
Equation~(\ref{eq:eq_of_state}), the equation of state.
%We have dropped the subscript b in this and the following two sections because we will be referring solely to baryonic quantities.  
The interpolation formula for some quantity $q$ is given by:
%
\begin{eqnarray}
q_j(x) & = &  q_{L,j} + \tilde{x}(\Delta q_j + q_{6,j}(1-\tilde{x})), \\
\tilde{x}      & \equiv & {x - x_{j-1/2} \over \Delta x_j}, \qquad
             x_{j-1/2} \leq x \leq x_{j+1/2}. \nonumber
\end{eqnarray}
%
This is Equation~(1.4) of CW84 (in the radial rather than mass coordinate as is appropriate for the direct rulerian approach). 
%The zones edges in the mass coordinates are $m_{j-1/2}$ and $m_{j+1/2}$ for the left and right edges, respectively.  
The quantities $q_{L,j}$, $\Delta q_j$,
and $q_{6,j}$ can be viewed simply as interpolation constants; however,
they also have more intuitive meanings.  For example, $q_{L,j}$ is the
value of $q$ at the left edge of zone j, while $\Delta q_j$ and $q_{6,j}$ are analogous to the slope and first order correction to the slope of $q$ (see \citet{1984JCoPh..54..174C} for a complete discussion):
\begin{equation}
\Delta q_j \equiv q_{R,j} - q_{L,j} \qquad 
q_{6,j}    \equiv 6\left[q_j - 1/2\left(q_{L,j} + q_{R,j}\right)\right].
\end{equation}

We have reduced the problem to finding $q_{L,j}$ and $q_{R,j}$.  While this
is simple in principle, it is complicated somewhat by the requirement that
these values be of sufficient accuracy and that the resulting distribution
be monotonic.  That is, no new maxima or minima are introduced.
The resulting formulae are straightforward but complicated and are not
reproduced here, but see Equations 1.7 to 1.10 of \citet{1984JCoPh..54..174C}.
We also optionally allow steepening as described in this reference.

@@

In the Eulerian approach, this is more complex because the Eulerian
zone boundary does not naturally correspond to one the three characteristic waves and so a more complicated approach is required.  First, the primary quantities ($p$, $\rho$, $v$ and $E$) are averaged over the domains corresponding to the three characteristics $C = u-c$, $u$ or $u+c$ (where $c$ is the sound speed in a cell):
\begin{eqnarray}
      q^{C+}_{j+1/2} & = & f_{j+1/2,L}(\Delta t C^n_j), \\
      q^{C-}_{j+1/2} & = & f_{j+1/2,R}(-\Delta t C^n_{j+1}), \nonumber
\end{eqnarray}
where $q^{C+}$ ($q^{C-}$) refers to the left (right) characteristic $C$.  The functions $f$ are just integrals
over the parabolic interpolation formula:
%
\begin{eqnarray}
f_{j+1/2,L}(y) & = &
       {1 \over y} \int^{x_{j+1/2}}_{x_{j+1/2}-y} q_j(x) dx, \\
f_{j+1/2,R}(y) & = &
       {1 \over y} \int^{x_{j+1/2}+y}_{x_{j+1/2}} q_{j+1}(x) dx. \nonumber
\end{eqnarray}
The linearized gas dynamics equations are then used to compute second-order correct left and right states that take into account the multiple wave families.  This process is described in \citet{1984JCoPh..54..174C} and we use their equations (3.6) and (3.7).

With these effective states, an approximation to the Riemann problem is found
with an iterative approach \citep[see][]{Woodward86}, producing estimates for 
$\overline{p}_{j\pm 1/2}$, $\overline{\rho}_{j\pm1/2}$, and $\overline{v}_{j\pm 1/2}$ that are third order
accurate in space and second order accurate in time.  These are then
used to solve the difference Equations (\ref{eq:mass_diff})--(\ref{eq:energy_diff}) for $\rho^{n+1}$, $v^{n+1}$, and $E^{n+1}$.

We include an optional diffusive flux (and flattening for the parabolic curves) that can improve the solution in some cases.  Our implementation follows that in the appendix of \citet{1984JCoPh..54..174C}.

In addition, as discussed earlier, the three dimensional scheme is achieved by operator splitting and repeating the above procedure in the other two orthogonal directions.  The transverse velocities and any additional passive quantities are naturally and easily added to this system (see Equation 3.6 of \citet{1984JCoPh..54..174C}).

We note that the acceleration required in Equation~(\ref{eq:momentum_diff}) is actually the
acceleration felt by the entire zone and not just at the zone center.
Therefore, it is possible to find the mass weighted average acceleration over the zone
by expanding the density and acceleration distributions and
retaining all terms up to second order in $\Delta x$:
%
\begin{equation}
g_j^{n+1/2} = 
       \frac{1}{2 a^{n+1/2} \delta x_j} \left[ 
             \phi_{j+1}^{n+1/2} 
           - \phi_{j-1}^{n+1/2} 
           + \frac{1}{12} \left(    \phi^{n+1/2}_{j+1} 
                                 - 2\phi^{n+1/2}_j 
                                 + \phi^{n+1/2}_{j-1} \right) 
                                   {\delta d_j \over d_j}
       \right].
       \label{eq:accel}
\end{equation}
Generally we find that the potential is so slowly varying that this change is unnecessary.

@@

The system described thus far works well for gravitating systems with
reasonable Mach numbers ($<100$) as long as the structures are well resolved.
This section and the next detail changes that are required
to correctly account for situations where one or both of these
requirements are not met.

Large, hypersonic bulk flows appear to be very common in cosmological
simulations and they present a problem because of the high ratio of
kinetic energy $E_k$ to gas internal energy $e$, which can reach as
high as $10^8$.  Inverted, we see that the internal energy consists of
an extremely small portion of the total energy.  The pressure then,
proportional to $E - E_{k}$, is the small difference between two large
numbers: a disastrous numerical situation.  This is not as large a
problem as it may at first appear because it only occurs when the
pressure is negligibly small.  Therefore, even if we suffer large
errors in the pressure distribution in these regions, the dynamics and
total energy budget of the flow will remain unaffected.  Nevertheless,
if the temperature distribution is required for other reasons
(e.g., for calculating radiative processes), a remedy is required.

To overcome this, we also solve the internal energy equation:
\begin{equation}
 \frac{\partial e}{\partial t} 
           + \frac{1}{a} \vecv \cdot \nabla e
%         = - \frac{3(\gamma -1)\dot{a}}{a} e
         = - 3 \frac{\dot{a}}{a} \frac{p}{\rho}
           - \frac{p}{a\rho} \nabla \cdot \vecv
\end{equation}
in comoving coordinates.  The structure is similar to the total energy
equation; the second term on the left hand side represents transport, while
the first term on the right is due to expansion of the coordinate
system.  
It is differenced (again, in Eulerian form without the expansion term) as,
%
\begin{equation}
\rho_j^{n+1} e_j^{n+1}  = 
       \rho_j^n e_j^n   +  \Delta t  \left(
            \frac{  \avp{\rho} \avp{v} \avp{e}  - \avm{\rho} \avm{v} \avm{e}} {\Delta x_j} \right)
            - \Delta t \ p_j^{n} \left( \frac{ \avp{v} - \avm{v}} {\Delta x_j} 
                      \right)
                  \label{eq:gasenergy_diff}
\end{equation}
%
Note that because of the structure of this equation, it is not in
flux-conservative form.  In particular, the pressure is evaluated at
the cell center.  Unfortunately, time-centering of this pressure has
proved difficult to do without generating large errors in the internal
energy and so we conservatively leave the pressure at the old time in
this difference equation.  This leads to spreading of shocks; however,
we note that this equation is only used in hypersonic flows.

@@

It should be stressed that as long as the parameter ${\eta}_1$ is small enough
the dual energy method {\it will have no dynamical effect}.  We use
${\eta}_1 = 10^{-3}$, which is consistent with the truncation error of the
scheme.  We are now free to select the method by which the internal energy
field variable $e$ is updated so that it will not become contaminated with
errors advected by the total energy formulation but still give the
correct distribution in shocked regions. 
Since we are concerned with the advection of errors, the selection
criterion must look at each cell's local neighbourhood.
In one dimension, this is done with,
\begin{equation}
e = \cases{ (E - \vecv^2/2),& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) > {\eta}_2$,\cr
            e,& $\rho(E - \vecv^2/2)/
    \max ({\rho}_{j-1}E_{j-1},{\rho}_j E_j,{\rho}_{j+1}E_{j+1}) < {\eta}_2$.\cr}
    \label{eq:dualselect}
\end{equation}
Thus, ${\eta}_2$ determines when the synchronization (of $e$ with $E$) occurs.
%is a measure of how much error we will allow to be passed
%from neighboring cells.  
Too high a value may mask relatively weak shocks, 
while spurious heating (via contamination) may occur if it is set too low.
After some experimentation,
we have chosen ${\eta}_2 = 0.08$, a somewhat conservative value.
This scheme is optional and is generally only required in large-scale 
cosmological simulations where the gas cools to the expansion of the universe
but large bulk flows develop due to the formation of structure.
